Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-04-01 paper/metamonads.tex
J's minor edits in O's sections

The essence of our approach is assembling the code of the program by
combining instances of various aspects together. In the case of
Gaussian Elimination, the aspects are pivot determination, swapping of
rows and columns, determinant tracking, etc. A primitive code
generator yields a code expression.  A combinator takes (complete and
well-typed) code expressions and combines them into a composite expression
(which is too statically guaranteed to be well-formed and well-typed).

We will be using MetaOcaml, which, as the instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them.

@@

(in the floating-point instance of that aspect, we would have
|.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
code expressions -- in other words, to construct future-stage
computations. The simplest combination is inlining:
\begin{code}
let plus x y = .<.~x + .~y>.
\end{code}
Here we see the second component of MetaOCaml -- escapes |.~|. They
let us perform an immediate code generating computation while we are
building the future-stage computation (code expression). In the
example above, the immediate computation |.~x| is trivial -- obtaining
the value of the variable |x|, which should be a code expression, and
inlining it. The following shows an example of using the simple
generators:

@@

Here, the immediate computation |gen .<x>. .<y>.| is less trivial --
it includes the evaluation of the function |gen|, which in turn
applies the function |plus|. The function receives code expressions
|.<x>.| and |.<y>.| as arguments. At the generating stage, we can
manipulate code expressions as values. The function |gen| returns a
code expression, which will be inlined in the place of the
escape. MetaOCaml can print out code expressions, so we can see the
final generated code:\\
  |.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|

We must point out that |gen x y| looks like a combinator; its body 
involves several applications of the function |plus|. When
we look at the generated code however, we see no traces of function
|plus| and no such administrative applications. Instead, we see the
fully inlined expression. These administative applications are done at
the code generation stage.

@@

and use it to generate code that operates on integers, floating point
numbers or booleans -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields the function on |float|s, whereas
the latter expression is the code expression for a bollean function.

We can see the separation of concerns, namely of that for domain
operations.

@@

We notice two occurrences of |plus y one|. Depending on the
implementation of |plus| that may be quite a complex computation, and
so we would rather not do it twice. We may be tempted to rely on the
compiler's common-subexpression elimination optimization. We would
rather not relied on a ``sufficiently smart compiler'': when the
generated code is very complex, the compiler may overlook common
subexpressions. The subexpressions may occur in such an imperative
context that the compiler might not be able to determine if lifting
them is sound. So, being conservative, the optimizer will leave the
duplicates as they are. We may attempt to write

@@

However, the result of |param_code1 plus one|, which prints as\\
|.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|,
still exhibits duplicate sub-expressions. Our |let|-insertion
optimization saved the computation on the generating stage. 
We need a combinator that inserts the |let| expression in the
generat\emph{ed} code. We need a combinator |letgen| to be used
as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
yielding the code like |.<let t = y + 1 in t + (x + t)>.|
But that seems impossible because |letgen exp| has to generate
the expression\\|.<let t = exp in body>.| but |letgen| does not have
the |body| yet. The body needs that temporary identifier |.<t>.| that
is supposed to be the result of |letgen| itself. 
Certainly |letgen| cannot generate only a part of a let-expression,
without the |body|.  All generated expressions in
MetaOCaml are well-formed and complete.

The key is to use continuation-passing style (CPS). It was first
pointed out by \cite{Bondorf:92} in the context
of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
generation.

@@

Comparing the code that did let-insertion at the generating stage\\
|let ce = (plus y one) in  plus ce (plus x ce)|\\
with the corresponding code inserting let at the generated code stage\\
|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
clearly shows the difference between the direct-style and CPS code.
What was |let ce = init in ...| in direct style became
|init' (fun ce -> ...)| in CPS. For one thing, |let| became
inverted. For another, what used to be an expression that yields
value, |init|, now became an expression that takes an extra argument,
the continuation, and invokes it. The differences look negligible in
the above example. In larger expressions with many let-forms, the
number of paranetheses around |fun| increases, the need to add and
then invoke the |k|, continuation argument, become annoying. The
inconvenience is great enough for some people to explicitly avoid CPS
or claim that numerical programmers (our users) cannot or will not
program in CPS. Clearly a better notation is needed.

The |do|-notation of Haskell (cite report?) shows that it is possible
to write essentially CPS code in the conventional-looking style. The
|do| notation is the notation for writing monadic code (cite Moggi?).
The benefit of monadic notation is not that it can represent CPS (cite
Filinski), but it helps in composability by offering to add different
layers of effects (state, exception, non-determinism, etc) to the
basic monad (cite Hudak, Peyton-Jones).

A monad (cite Moggi?) is an abstract datatype representing
computations that yield a value and may have an \emph{effect}.
The datatype must have at least two operations, |return| to build
trivial effect-less computations and |bind| for combining
computations. These operations must satisfy \emph{monadic laws}:
|return| being the left and the right unit of |bind| and |bind| being
associative. Figure XXX defines the monad in the present paper
and shows its implementation.

\begin{figure*}
\begin{code}
type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
let ret a = fun s k -> k s a
let bind a f = fun s k -> a s (fun s' b -> f b s' k)
let fetch s k = k s s
let store v s k = k (v::s) ()

@@

let lif test th el = ret .< if .~test then .~th else .~el >.
let ifM test th el = fun s k ->
  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
\end{code}
\caption{Our monad}
\end{figure*}

@@

In Figure XXX, the monad is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state, the value (of the type |v|) and
yields the answer of the type |w|.  The monad is polymorphic over all
these three type parameters. Other implementations are
possible. Except the code in Figure XXX, the rest of our code treats
the monad as a truly abstract data type. The implementation of basic
monadic operations |ret| and |bind| are conventional. It is easy to
see that the monadic laws are satisfied. Other monadic operations
construct computations that do have effects. Operations |fetch| and
|store v| construct computations that read and write the state. In our
case the state is a list (of polymorphic variants), which models an
open discriminated union, as we shall see later.

The operation |retN a| is the let-insertion operation (whose simpler
version was called |letgen| before). It is the first computation with
a control effect: indeed, the result of |retN a| is \emph{not} the
result of invoking its continuation |k|. Rather, its result is a |let|
code expression. Mention similarity with reset?

Finally, |runM| runs our monad, that is, performs the computation of
the monad and get the result, which in our case is the code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as

@@

That does not seem like the improvement. But here we can introduce the
do notation (Appendix A), which is patterned after the do-notation of Haskell.

@@

The function |param_code4| is totally equivalent to |param_code3| --
the |mdo|-notation is just the syntactic sugar. In fact, the camlp4
preprocessor (with our extension) will transform the |param_code4|
into |param_code3|. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

\oleg{Should we rename lif into ifL?}
We can write operations that generate code other than let-statements,
e.g., conditionals: |lif| (Figure XXX). The function |lif|, albeit
straightforward, is not as general as we wish: its argument are
already generated pieces of code rather than monadic values. So, we
may want to ``lift it'', see |ifM'|:

\begin{code}
let ifM' test th el = mdo {
  testc <-- test;
  thc   <-- th;
  elc   <-- el;
  lif testc thc elc}
\end{code}

\omitnow{
\begin{code}
let ifM' test th el = bind test (fun testc ->
                      bind th   (fun thc ->
                      bind el   (fun elc ->
		      lif testc thc elc)));;
\end{code}
}

We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
|liftM2|, |liftM3| of Haskell) to make such a lifting generic.
We soon notice the need in another |ifM| function, with tha same
interface (see ifM in Fig. XXX). The difference between them is
apparent from the following:

\begin{code}
let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
                           (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
;;
.<fun a_1 ->
   fun i_2 ->
    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
\end{code}

whereas if we replace |ifM'| with |ifM|, the result will be

\omitnow{
\begin{code}
let gen a i = ifM (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
                           (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
;;
\end{code}
}
\begin{code}
.<fun a_1 ->
   fun i_2 ->
    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
\end{code}

The difference is apparent: in the code with |ifM'|, the let-insertion
happend \emph{before} the if-expression, that is, before the test that
the index |i| is positive. That is, if |i| turned out
negative, |a.(i)| will generate the out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that the code generation (such as the one in |retN|) is 
truly an effect and we have to be clear about the sequencing of
effects, when generating control constrcutions such as conditionals.
The form |ifM| handles the such effects correctly. In our code, we
need similar operators for other Ocaml control forms: for generating
case-matching statements and for- and while-loops. To make writing
such code general and independent of the implementation of the monad,
we define a function XXX. As the name implies, we do use control
effect implemented via essentially delimited continuations (cite?).

>>>>

The essence of our approach is assembling the code of the program by
combining instances of various aspects together. In the case of
Gaussian Elimination, the algorithmic aspects are pivot determination, 
swapping of
rows and columns, determinant tracking, etc. A primitive code
generator yields a code expression.  A combinator takes (complete and
well-typed) code expressions and combines them into a composite expression
(which is also statically guaranteed to be well-formed and well-typed).

We will be using MetaOcaml which, as an instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them.

@@

In the floating-point instance of that aspect, we would have
|.<1.0>.| instead. We use MetaOCaml brackets |.<...>.| to generate
code expressions -- in other words, to construct future-stage
computations. The simplest combination is inlining:
\begin{code}
let plus x y = .<.~x + .~y>.
\end{code}
Here we see the second component of MetaOCaml -- escapes |.~|. They
let us perform an immediate code generating computation while we are
building the future-stage computation (code expression). In the
example above, the immediate computation |.~x| is trivial -- obtaining
the value of the variable |x|, which should be a code expression, and
inlining it. The following shows an example of using the simple
generators:

@@

Here, the immediate computation |gen .<x>. .<y>.| is less trivial --
it includes the evaluation of the function |gen|, which in turn
applies the function |plus|. The function receives code expressions
|.<x>.| and |.<y>.| as arguments. At the generating stage, we can
manipulate code expressions as (opaque) values. The function |gen| returns a
code expression, which will be inlined in the place of the
escape. MetaOCaml can print out code expressions, so we can see the
final generated code:\\
  |.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|

We must point out that |gen x y| looks like a combinator; its body 
involves several applications of the function |plus|. When
we look at the generated code however, we see no traces of the function
|plus| and no such administrative applications. Instead, we see the
fully inlined expression. These administative applications are done at
the code generation stage.

@@

\noindent and use it to generate code that operates on integers, floating point
numbers or booleans -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields the function on |float|s, whereas
the latter expression is the code expression for a boolean function.

This clearly shows the separation of concerns, namely of that for domain
operations.

@@

We notice two occurrences of |plus y one|. Depending on the
implementation of |plus| that may be quite a complex computation, and
so we would rather not do it twice. We may be tempted to rely on the
compiler's common-subexpression elimination optimization. We would
rather not rely on a ``sufficiently smart compiler'': when the
generated code is very complex, the compiler may overlook common
subexpressions.  Or the subexpressions may occur in an imperative
context where the compiler might not be able to determine if lifting
them is sound. So, being conservative, the optimizer will leave the
duplicates as they are. We may attempt to write

@@

However, the result of |param_code1 plus one|, which prints as\\
|.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|,
still exhibits duplicate sub-expressions. Our |let|-insertion
optimization saved the computation at the generating stage. 
We need a combinator that inserts the |let| expression in the
generat\emph{ed} code. We need a combinator |letgen| to be used
as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
yielding the code like |.<let t = y + 1 in t + (x + t)>.|
But that seems impossible because |letgen exp| has to generate
the expression\\|.<let t = exp in body>.| although |letgen| does not have
the |body| yet. The body needs a temporary identifier |.<t>.| that
is supposed to be the result of |letgen| itself. 
Certainly |letgen| cannot generate only part of a let-expression,
without the |body|,  as all generated expressions in
MetaOCaml are well-formed and complete.

The key is to use continuation-passing style (CPS). Its benefits were first
pointed out by \cite{Bondorf:92} in the context
of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
generation.

@@

Comparing the code that did let-insertion at the generating stage\\
|let ce = (plus y one) in  plus ce (plus x ce)|\\
with the corresponding code inserting let at the generated code stage\\
|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
clearly shows the difference between the direct-style and CPS code.
What was |let ce = init in ...| in direct style became
|init' (fun ce -> ...)| in CPS. For one thing, |let| became
inverted. For another, what used to be an expression that yields
a value, |init|, now became an expression that takes an extra argument,
the continuation, and invokes it. The differences look negligible in
the above example. In larger expressions with many let-forms, the
number of parentheses around |fun| increases, the need to add and
then invoke the |k| continuation argument become increasingly annoying. The
inconvenience is great enough for some people to explicitly avoid CPS
or claim that numerical programmers (our users) cannot or will not
program in CPS. Clearly a better notation is needed.

The |do|-notation of Haskell \cite{Haskell98Report} shows that it is possible
to write essentially CPS code in conventional-looking style. The
|do| notation is the notation for writing monadic code (cite Moggi?).
The benefit of monadic notation is not that it can represent CPS (cite
Filinski), but it helps in composability by offering to add different
layers of effects (state, exception, non-determinism, etc) to the
basic monad (cite Hudak, Peyton-Jones) in a controlled way.

A monad (cite Moggi?) is an abstract datatype representing
computations that yield a value and may have an \emph{effect}.
The datatype must have at least two operations, |return| to build
trivial effect-less computations and |bind| for combining
computations. These operations must satisfy \emph{monadic laws}:
|return| being the left and the right unit of |bind| and |bind| being
associative. Figure~\ref{ourmonad} defines the monad used throughout
the present paper and shows its implementation.

\begin{figure*}\label{ourmonad}
\begin{code}
type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
let ret a = fun s k -> k s a
let bind a f = fun s k -> a s (fun s' b -> f b s' k)
let fetch s k = k s s
let store v s k = k (v::s) ()

@@

let ifL test th el = ret .< if .~test then .~th else .~el >.
let ifM test th el = fun s k ->
  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
\end{code}
\caption{Our monad}
\end{figure*}

@@

In Figure~\ref{ourmonad}, the monad is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state, the value (of the type |v|) and
yields the answer of the type |w|.  The monad is polymorphic over all
these three type parameters.  Other implementations are
possible. Except for the code in Figure~\ref{ourmonad}, the rest of our code
treats the monad as a truly abstract data type. The implementation of basic
monadic operations |ret| and |bind| are conventional. It is easy to
see that the monadic laws are satisfied.  Other monadic operations
construct computations that do have effects.  Operations |fetch| and
|store v| construct computations that read and write the state. In our
case the state is a list (of polymorphic variants), which models an
open discriminated union, as we shall see later.

The operation |retN a| is the let-insertion operation, whose simpler
version we called |letgen| earlier. It is the first computation with
a control effect: indeed, the result of |retN a| is \emph{not} the
result of invoking its continuation |k|. Rather, its result is a |let|
code expression. \oleg{Mention similarity with reset?}
\jacques{yes}

Finally, |runM| runs our monad, that is, performs the computation of
the monad and gets the result, which in our case is the code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as

@@

That does not seem like much of an improvement. But here we can introduce the
do notation (Appendix A), which is patterned after the do-notation of Haskell.

@@

The function |param_code4| is totally equivalent to |param_code3| --
the |mdo|-notation is just syntactic sugar. In fact, the camlp4
preprocessor (with our extension) will transform |param_code4|
into exactly |param_code3|. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its argument are
already generated pieces of code rather than monadic values. So, we
may want to ``lift it'':
\begin{code}
let ifM' test th el = mdo {
  testc <-- test;
  thc   <-- th;
  elc   <-- el;
  ifL testc thc elc}
\end{code}

\noindent We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
|liftM2|, |liftM3| of Haskell) to make such a lifting generic.
We soon notice the need for another |ifM| function, with the same
interface (see ifM in Figure~\ref{ourmonad}). The difference between them is
apparent from the following:
\begin{code}
let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
                           (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
;;
.<fun a_1 -> fun i_2 ->
    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
\end{code}
\noindent whereas if we replace |ifM'| with |ifM|, the result will be
\begin{code}
.<fun a_1 -> fun i_2 ->
    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
\end{code}

The difference is apparent: in the code with |ifM'|, the let-insertion
happend \emph{before} the if-expression, that is, before the test that
the index |i| is positive. That is, if |i| turned out
negative, |a.(i)| will generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that the code generation (such as the one in |retN|) is 
truly an effect and we have to be clear about the sequencing of
effects, when generating control constructions such as conditionals.
The form |ifM| handles the such effects correctly. In our code, we
need similar operators for other Ocaml control forms: for generating
case-matching statements and for- and while-loops. To make writing
such code general and independent of the implementation of the monad,
we define a function XXX. As the name implies, we do use control
effect implemented via essentially delimited continuations (cite?).
\jacques{yes!}
